---
title: "Gapminder Data Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 8,
  fig.height = 5
)
pkgs <- c("ggplot2","dplyr","gapminder","scales","viridis","tidyr","forcats","zoo")
for(pkg in pkgs) if(!requireNamespace(pkg, quietly=TRUE)) install.packages(pkg)
lapply(pkgs, library, character.only=TRUE)
````

# 1. Introduction

This analysis explores global trends in life expectancy, GDP per capita, and population from 1952 to 2007 using the Gapminder dataset. We will produce 12 charts and fit a regression model.

# 2. Data Loading & Preparation

```{r load-data}
data("gapminder")
gm <- gapminder

latest <- gm %>% 
  filter(year == max(year))

global_trends <- gm %>%
  group_by(year) %>%
  summarize(
    avg_lifeExp   = mean(lifeExp),
    avg_gdpPercap = mean(gdpPercap),
    total_pop     = sum(pop),
    .groups = "drop"
  )
```

# 3. Global Trends

## 3.1 Global Average Life Expectancy

```{r global-lifeExp}
p1 <- ggplot(global_trends, aes(year, avg_lifeExp)) +
  geom_line(color="steelblue", size=1) +
  labs(
    title = "Global Average Life Expectancy Over Time",
    x = "Year", y = "Life Expectancy (years)"
  ) +
  theme_minimal()
print(p1)
```

## 3.2 Global Average GDP per Capita

```{r global-gdp}
p2 <- ggplot(global_trends, aes(year, avg_gdpPercap)) +
  geom_line(color="darkgreen", size=1) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Global Average GDP per Capita Over Time",
    x = "Year", y = "GDP per Capita (USD)"
  ) +
  theme_minimal()
print(p2)
```

## 3.3 Global Total Population

```{r global-pop}
p3 <- ggplot(global_trends, aes(year, total_pop/1e9)) +
  geom_line(color="purple", size=1) +
  labs(
    title = "Global Total Population Over Time",
    x = "Year", y = "Population (billions)"
  ) +
  theme_minimal()
print(p3)
```

# 4. Country-Level Analysis (2007)

## 4.1 Life Expectancy vs GDP per Capita

```{r scatter-lifespan-gdp}
p4 <- ggplot(latest, aes(gdpPercap, lifeExp)) +
  geom_point(alpha=0.7) +
  scale_x_log10(labels = scales::dollar_format()) +
  labs(
    title = "Life Expectancy vs GDP per Capita (2007)",
    x = "GDP per Capita (log scale)", y = "Life Expectancy (years)"
  ) +
  theme_minimal()
print(p4)
```

## 4.2 Regression: lifeExp \~ log(gdpPercap)

```{r regression}
model <- lm(lifeExp ~ log(gdpPercap), data = latest)
cat("Regression summary:\n")
print(summary(model))

p5 <- ggplot(latest, aes(log(gdpPercap), lifeExp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, color = "darkred") +
  labs(
    title = "Regression: Life Expectancy ~ log(GDP per Capita)",
    x = "log(GDP per Capita)", y = "Life Expectancy"
  ) +
  theme_minimal()
print(p5)
```

## 4.3 Top 10 Countries by Life Expectancy

```{r top10-lifeExp}
top10_le <- latest %>%
  arrange(desc(lifeExp)) %>%
  slice_head(n = 10)
p6 <- ggplot(top10_le, aes(fct_reorder(country, lifeExp), lifeExp)) +
  geom_col(fill = "darkslateblue") +
  coord_flip() +
  labs(
    title = "Top 10 Countries by Life Expectancy (2007)",
    x = NULL, y = "Life Expectancy (years)"
  ) +
  theme_minimal()
print(p6)
```

# 5. Continental Comparisons

## 5.1 Boxplot: Life Expectancy by Continent

```{r boxplot-continent}
p7 <- ggplot(latest, aes(continent, lifeExp, fill = continent)) +
  geom_boxplot() +
  labs(
    title = "Life Expectancy by Continent (2007)",
    x = "Continent", y = "Life Expectancy (years)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
print(p7)
```

## 5.2 Violin: GDP per Capita by Continent

```{r violin-gdp}
p8 <- ggplot(latest, aes(continent, gdpPercap, fill = continent)) +
  geom_violin() +
  scale_y_log10(labels = scales::dollar_format()) +
  labs(
    title = "GDP per Capita Distribution by Continent (2007)",
    x = "Continent", y = "GDP per Capita (log scale)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
print(p8)
```

# 6. Trends by Continent

## 6.1 Life Expectancy Over Time

```{r trend-by-continent}
p9 <- ggplot(gm, aes(year, lifeExp, color = continent)) +
  geom_line(stat="summary", fun="mean", size=1) +
  labs(
    title = "Average Life Expectancy Over Time by Continent",
    x = "Year", y = "Life Expectancy (years)", color = "Continent"
  ) +
  theme_minimal()
print(p9)
```

# 7. Distribution & Heatmap

## 7.1 Density: Life Expectancy (2007)

```{r density-lifeExp}
p10 <- ggplot(latest, aes(lifeExp, fill = continent)) +
  geom_density(alpha = 0.6) +
  labs(
    title = "Density of Life Expectancy by Continent (2007)",
    x = "Life Expectancy (years)", y = "Density"
  ) +
  theme_minimal()
print(p10)
```

## 7.2 Heatmap: Life Expectancy by Year & Continent

```{r heatmap-lifeExp}
heat_data <- gm %>%
  group_by(continent, year) %>%
  summarize(avg_le = mean(lifeExp), .groups="drop")
p11 <- ggplot(heat_data, aes(year, continent, fill = avg_le)) +
  geom_tile(color = "white") +
  scale_fill_viridis(name = "Avg Life Exp") +
  labs(
    title = "Average Life Expectancy by Year & Continent",
    x = "Year", y = "Continent"
  ) +
  theme_minimal()
print(p11)
```

# 8. Population Bubble Plot

```{r bubble-plot}
p12 <- ggplot(latest, aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.6) +
  scale_x_log10(labels = scales::dollar_format()) +
  scale_size_continuous(
    labels = scales::comma_format(scale = 1e-6),
    name   = "Population\n(millions)"
  ) +
  labs(
    title = "Life Expectancy vs GDP per Capita (2007) Sized by Population",
    x     = "GDP per Capita (log scale)",
    y     = "Life Expectancy"
  ) +
  theme_minimal()
print(p12)
```

# 9. Conclusions

This analysis highlights:

* Steady increases in global life expectancy and GDP per capita.
* Strong positive correlation between wealth and longevity.
* Significant continental differences in health and economic outcomes.

Further extensions could examine fertility, child mortality, or educational attainment as additional dimensions.
